### 2018-05-15 Final Paper/paper.tex
Added Stephanie's changes
fixed up the paper a little

\section{Introduction}
The study of circadian clocks within organisms has been a growing field in systems biology.  We specifically looked at the oscillations of Per2 and BMal1 proteins in cells damaged by cancer.  Michelle Farkas, from the University of Massachusetts Amherst Department of Chemistry, provided us data from cancer-damaged cells that were starved for 24 hours, 12 hours, and not starved at all.  It's important to understand the period of these oscillations as knowing the period plays into understanding the mechanics of the larger cellular systems.  However, due to the nature of the data there were no consistently accurate automated peak detection system in place.  The data has a couple of idiosyncrasies that made automated analysis harder - the oscillations are damped, there was a downwards trend in all of the datasets, and there was a lot of noise in the data. Furthermore, most period finding algorithms are designed to work for data that has a constant period, but we found that this data had a period that varied with time.
Moving forward with the task of determining the period we proceeded with several initial steps.  This included analyzing smoothing techniques as we needed to smooth the data while ensuring that it remained as true to the actual data as possible through manipulations and derivatives.   Additionally, the use of genetic algorithms to create a multiple cosines function for obtaining the periodicity was explored.  The classic use of derivatives for finding peaks to calculate the period was also investigated for determining the best automated period detection tool.

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/FourFit_GA_match"}.png}
	\caption{A plot of artificially generated data that has then been fit to using the mFourFit equation.}
	\label{fig:mFourFit_ex}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/corrected_poly"}.png}
	\caption{Using the third-degree polynomial fit to find the downwards trend of our data, and correcting the data based on that downwards trend.}
	\label{fig:correcting_with_poly}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/first_drv_fail"}.png}
	\caption{Using the first derivative to find peaks.}
	\label{fig:first_drv_fail}
\end{figure}

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/third_drv_fail"}.png}
	\caption{Using the third derivative to find peaks.}
	\label{fig:third_drv_fail}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/Final process"}.png}
	\caption{Our final peak and valley detection process. This process worked consistently for every dataset that we tried it on.}
	\label{fig:pv_det_final}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/corrected_poly"}.png}
	\caption{A plot of the found period over time of one of our datasets. We also plotted the third-degree polynomial fit of the found period data in order to get a continuous function of the period over time.}
	\label{fig:per_over_time}
\end{figure}

>>>>

\section{Introduction}
The study of circadian clocks in organisms has been an important endeavor for over 50 years. We examined bio-luminescence recordings from human breast cancer model cells with two circadian reporters: Per2 and Bmal1 (Michelle Farkas, Dept. of Chemistry, University of Massachusetts Amherst). It's important to understand the period of these oscillations as knowing the period plays into understanding the mechanics of the larger cellular systems.  However, due to the nature of the data there were no consistently accurate automated peak detection system in place.  The data has a couple of idiosyncrasies that made automated analysis harder - the oscillations are damped, there was a downwards trend in all of the datasets, and there was a lot of noise in the data. Furthermore, most period finding algorithms are designed to work for data that has a constant period, but we found that this data had a period that varied with time.\\
Moving forward with the task of determining the period we proceeded with several initial steps.  This included analyzing smoothing techniques as we needed to smooth the data while ensuring that it remained as true to the actual data as possible through manipulations and derivatives.   Additionally, the use of genetic algorithms to create a multiple cosines function for obtaining the periodicity was explored.  The classic use of derivatives for finding peaks to calculate the period was also investigated for determining the best automated period detection tool.

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/FourFit_GA_match"}.png}
	\caption{A plot of artificially generated data that has then been fit to using the mFourFit equation using a genetic algorithm. Notice that although the fit doesn't perfectly match the artificial data, it does catch every oscillation that it makes (i.e. has a peak everywhere there is a peak and a valley everywhere there is a valley).}
	\label{fig:mFourFit_ex}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/corrected_poly"}.png}
	\caption{This plot demonstrates the process of using the third-degree polynomial fit to find the downwards trend of our data. We then correct the data based on that downwards trend. We do this by adding the inverse of the trend to our original data.}
	\label{fig:correcting_with_poly}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/first_drv_fail"}.png}
	\caption{An example of using the first derivative to find peaks and valleys. It is clear that although this finds a good number of peaks and valleys, there are some that were not found. In this case, we notice a bump at about 3 days in, where you can see a valley and then a peak, but the code doesn't catch it.}
	\label{fig:first_drv_fail}
\end{figure}

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/third_drv_fail"}.png}
	\caption{An example of using the third derivative to find peaks and valleys. It is clear that although this doesn't miss any peaks or valleys, it is overly sensitive. It not only finds false peaks (around 1.4 days in), but also finds multiple peaks in one spot (for example, at around 2 days in).}
	\label{fig:third_drv_fail}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/Final process"}.png}
	\caption{A diagram of our final peak and valley detection process. This process in total produces a plot of discrete periods, and a function of time that gives us the period. Unlike any of the other methods we tried, this worked equally well for all of our datasets, making it robust enough to be trusted for automation.}
	\label{fig:pv_det_final}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\linewidth]{{"plots/per_over_time"}.png}
	\caption{A plot of the found period over time of one of our datasets. We also plotted the third-degree polynomial fit of the found period data in order to get a continuous function of the period over time.}
	\label{fig:per_over_time}
\end{figure}
